Synthesizing Quotient Lenses

### 2017-10-18 papers/QuotientLensSynth/main.tex
Pass through the intro

In Boomerang, equivalence relations are expressed using {\em canonizers}
which are functions that map members of a regular language to their canonical
representation modulo an equivalence relation. Therefore, in order to give a
specification for the data, the programmer often has to write a
regular expression which matches all possible representations of the data, and
then provide a set of canonizers which does the job of quotienting the data.

For example, suppose that the programmer wants to
derive the \bibtex{} to EndNote transformation in such a way that the program
allows for whitespace to occur arbitrarily between the words in the author and
title fields, ignores the order in which the author and title fields occur in
each record, and allows for arbitrary amounts of whitespace in between the
author and title fields. These rules allow for these two \bibtex{} citation
records to be considered the same,

\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines},
}

@Book{conway,

    Title = {Regular Algebra 
            and Finite Machines},

    Author = "J. H. Conway",
}
\end{verbatim}

\noindent with the EndNote records having a similar equivalence relation defined
on them.

The code sample in Figure~\ref{fig:example-canonizers} shows how the programmer
might specify these equivalence relations on the \bibtex{} records in Boomerang
using canonizers.

\begin{figure}[t]

\begin{lstlisting}
let project (r : regexp) (s : string) = canonizer_of_lens ((del r) . (ins s))
let wsp : regexp = [ \n\t\r]$^+$
let wsp_to_space : canonizer = project wsp " "
let valid_name : canonizer = 
    name | (name . wsp_to_space . (name .wsp_to_space)$^*$ .name)
let wsp_to_and : canonizer = project (wsp . "and" . wsp) " and "
let bib_names : canonizer =  
    valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author : canonizer = "author = \"" . bib_names . "\""
let label : regexp = [a-zA-Z0-9]$^+$
let title : canonizer = 
    word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title : canonizer = "title = {" . title . "}"
let wsp_to_comma : canonizer = project ("," . wsp) ",\n"
let first_perm : canonizer = bib_author . wsp_to_comma . bib_title
let second_perm : canonizer = bib_title . wsp_to_comma . bib_author
let bibtex : canonizer = 
    "@book{" . label. wsp_to_comma . (first_perm | second_perm) . "}"
\end{lstlisting}

  \caption{Boomerang canonizers for the \bibtex{}--EndnNote equivalence
    relations.}
  \label{fig:example-canonizers}
\end{figure}

Using the in-built \cd{canonizer_of_lens} function which extracts a canonizer
from a lens, it is easy to define an equivalence relation that sends all
members of a regular language to a canonical representative. For example
the expression 
\begin{lstlisting}
let wsp_to_space : canonizer = canonizer_of_lens ((del wsp) . (ins " "))
\end{lstlisting}
\noindent defines a canonizer which sends arbitrary occurences of whitespace
characters to a single space. However, a problem arises when trying to specify
the equivalence relation which nests this projection relation with the relation
that considers all permutations of the \bibtex{} fields equivalent.

Specifically, it becomes necessary to specify each of the permutations
seperately, since Boomerang's in-built permutation combinator does not account
for separators in between the regular expressions being permuted:

@@

One possible solution is to permute the alternative canonizers
\cd{bib_author.wsp_to_comma} and \cd{bib_title.wsp_to_comma},
but Boomerang's type-checker will reject this expression since the permutation
combinator accepts regular expressions rather than canonizers. In any case, the
regular expression
\begin{lstlisting}
bib_author . wsp_to_comma . bib_title . wsp_to_comma | 
bib_title . wsp_to_comma . bib_author . wsp_to_comma
\end{lstlisting}
is not equivalent to,
\begin{lstlisting}
bib_author . wsp_to_comma . bib_title  | 
bib_title . wsp_to_comma . bib_author
\end{lstlisting}
thus the programmer would have to write more lenses and canonizers to correct
this. Therefore, in order to utilize this trick, the programmer might choose to
specify all the regular expressions first (including the whole regular
expressions canonized by the \cd{bib_author} and \cd{bib_title}
canonizers), and then the canonizers separately. Alternatively, he/she may stick
to using canonizers and then liberally make use of the built-in
functions \cd{uncanonized_type} and \cd{canonized_type} which return the
regular expression matching all the data and the regular expression matching
the canonical representatives respectively.

@@

  \caption{QRE specification of \bibtex{} records.}
  \label{fig:example-qre}
\end{figure}

The expression \cd{project [ \n\t\r]+ -> " "} is a QRE that matches any
number of whitespace characters, and then considers all strings of whitespace
characters as equivalent with a single space as the canonical representative. 
Another QRE in the example above is the

@@

While QREs kill two birds with one stone by specifying the
data and the equivalence relation on the data in one expression, he/she
still has to manually program the underlying lens which maps the equivalence
classes of the source data to the equivalence classes of the view data, a task
which is itself tedious and error-prone as the code snippet below illustrates:

@@

Therefore, to address this challenge, we demonstrate how to {\em synthesize}
quotient lenses from a pair of QREs and a set of input-output example pairs.
Our key insight is to reduce the problem of synthesizing QRE lenses to the
problem of synthesizing {\em bijective lenses}, a particular subset of the
lenses that can be expressed in Boomerang. This is a problem which we solved in
our prior work ~\cite{popl18}.

@@

One potential pitfall with QRE lenses is that the combinators which
we define on them may not preserve this normal form. For example, the
composition $q' \circ q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens which has the canonical normal form of a bijective lens with
canonizers a the edges. Our main technical contribution in this paper is a
proof that QRE lenses are closed under left quotienting, right quotienting,
composition, and regular combinators. This proof will allow justify our
approach for synthesizing quotient lenses so that the programmer can simply
type

@@

We have already encountered the \verb!project R $\to$ s! QRE which sends all
elements of the language described by the regular expression $R$, and the

>>>>

\bcp{I'm nervous about making the concept of canonizer so central, since it
  is not going to be much of a focus for the rest of the paper.  In general,
it feels like the intro is explaining too much.  We need to find a more
streamlined path through the ideas!}
In Boomerang, equivalence relations are expressed using {\em canonizers}
which are functions that map members of a regular language to their canonical
representation modulo an equivalence relation. Therefore, in order to give a
specification for the data, the programmer often has to write a
regular expression which matches all possible representations of the data, and
then provide a canonizer that does the job of quotienting the data.

For example, suppose that the programmer wants to
derive the \bibtex{} to EndNote transformation in such a way that the program
allows for whitespace to occur arbitrarily between the words in the author and
title fields, ignores the order in which the author and title fields occur in
each record, and allows for arbitrary amounts of whitespace in between the
author and title fields. These rules allow for the \bibtex{} citations
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }
\end{verbatim}
and
\begin{verbatim}
    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "J. H. Conway",     }
\end{verbatim}
to be considered the same (with the EndNote records having a similar
equivalence relation defined on them).

\begin{figure}[t]
\begin{lstlisting}
let project (r : regexp) (s : string) = canonizer_of_lens ((del r) . (ins s))
let wsp : regexp = [ \n\t\r]$^+$
let wsp_to_space : canonizer = project wsp " "
let valid_name : canonizer = 
    name | (name . wsp_to_space . (name .wsp_to_space)$^*$ .name)
let wsp_to_and : canonizer = project (wsp . "and" . wsp) " and "
let bib_names : canonizer =  
    valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author : canonizer = "author = \"" . bib_names . "\""
let label : regexp = [a-zA-Z0-9]$^+$
let title : canonizer = 
    word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title : canonizer = "title = {" . title . "}"
let wsp_to_comma : canonizer = project ("," . wsp) ",\n"
let first_perm : canonizer = bib_author . wsp_to_comma . bib_title
let second_perm : canonizer = bib_title . wsp_to_comma . bib_author
let bibtex : canonizer = 
    "@book{" . label. wsp_to_comma . (first_perm | second_perm) . "}"
\end{lstlisting}

  \caption{Boomerang canonizers for the \bibtex{}--EndNote equivalence
    relations.}
  \label{fig:example-canonizers}
\end{figure}
The code sample in Figure~\ref{fig:example-canonizers} shows how the programmer
might specify these equivalence relations on the \bibtex{} records in Boomerang
using canonizers.
%
Using the in-built \cd{canonizer_of_lens} function which extracts a canonizer
from a lens, it is easy to define an equivalence relation that sends all
members of a regular language to a canonical representative.\bcp{Oh dear.
  The \cd{canonizer_of_lens} thing is really kind of a hack---we're using
  the fancy/specialized language of lenses to write what could really be a
  simple program in an arbitrary language (and then throwing away half of the
  lens, when the whole reason we give for having lenses in the first place
  is because we want both directions).  As an implementation trick, it was a
good idea.  But trying to get readers to understand it here does not seem
like a good idea.} For example
the expression 
\begin{lstlisting}
let wsp_to_space : canonizer = canonizer_of_lens ((del wsp) . (ins " "))
\end{lstlisting}
\noindent defines a canonizer which sends arbitrary occurences of whitespace
characters to a single space\bcp{Going into this level of detail encourages
readers to actually try to understand the details, but we don't actually
give enough details to allow them to do that!}. However, a problem arises
when trying to specify 
the equivalence relation which nests this projection relation with the relation
that considers all permutations of the \bibtex{} fields equivalent.

Specifically, it becomes necessary to specify each of the permutations
seperately, since Boomerang's in-built permutation combinator does not account
for separators in between the regular expressions being permuted:\bcp{Why
  couldn't we just enrich Boomerang's permutation combinator to do this?}

@@

One possible solution is to permute the alternative canonizers
\cd{bib_author.wsp_to_comma} and \cd{bib_title.wsp_to_comma},
but Boomerang's type-checker will reject this expression since the permutation
combinator accepts regular expressions rather than canonizers\bcp{It feels
  like we're lost, here, in details of exactly what Boomerang does and
  doesn't do.  This doesn't seem especially relevant to quotient lenses in
  general, or to what we're doing in this paper.}. In any case, the
regular expression
\begin{lstlisting}
bib_author . wsp_to_comma . bib_title . wsp_to_comma | 
bib_title . wsp_to_comma . bib_author . wsp_to_comma
\end{lstlisting}
is not equivalent to,
\begin{lstlisting}
bib_author . wsp_to_comma . bib_title  | 
bib_title . wsp_to_comma . bib_author
\end{lstlisting}
thus the programmer would have to write more lenses and canonizers to correct
this. Therefore, in order to utilize this trick, the programmer might choose to
specify all the regular expressions first (including the whole regular
expressions canonized by the \cd{bib_author} and \cd{bib_title}
canonizers), and then the canonizers separately. Alternatively, he/she may stick
to using canonizers and then liberally make use of the built-in
functions \cd{uncanonized_type} and \cd{canonized_type} which return the
regular expression matching all the data and the regular expression matching
the canonical representatives respectively.

@@

\caption{QRE specification of \bibtex{} records.  \bcp{I'm afraid this pair
    of examples is not making the point we intended.  I thought we were
    hoping to just say ``Here's the old Boomerang version; here's the QRE
    version; look how much smaller and simpler the second one is.''  But the
    second one {\em isn't} actually much smaller and, at least on first
    impression, doesn't look simpler.}}
  \label{fig:example-qre}
\end{figure}

The expression \bcp{mangled---did you proofread this?}\cd{project [
  \n\t\r]+ -> " "} is a QRE that matches any 
number of whitespace characters, and then considers all strings of whitespace
characters as equivalent with a single space as the canonical representative. 
Another QRE in the example above is the

@@

While QREs kill two birds with one stone by specifying the
data and the equivalence relation on the data in one expression,
he/she\bcp{Awkward: just use ``she'' (or find a way to alternate pronouns,
  if that doesn't create confusion)}
still has to manually program the underlying lens which maps the equivalence
classes of the source data to the equivalence classes of the view data, a task
which is itself tedious and error-prone as the code snippet below
illustrates:\bcp{If we added this to the old-Boomerang figure it might make
  a better case for size and simplicity!}

@@

Therefore, to address this challenge, we demonstrate how to {\em synthesize}
quotient lenses from a pair of QREs and a set of input-output example pairs.
Our key insight is to reduce the problem of synthesizing QRE lenses to the
problem of synthesizing {\em bijective lenses}, a particular subset of the
lenses that can be expressed in Boomerang.\bcp{As stated, this is
  misleading: the reader might have the impression that we somehow reduce
  the problem of synthesizing {\em asymmetric} quotient lenses to the
  problem of synthesizing bijective ordinary lenses!} This is a problem
which we solved in our prior work ~\cite{popl18}.

@@

One potential pitfall with QRE lenses is that the combinators which
we define on them may not preserve this normal form. For example, the
composition $q' \circ q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens which has the canonical normal form of a bijective lens with
canonizers a the edges. Our main technical contribution in this paper is a
proof that QRE lenses are closed under left quotienting, right quotienting,
composition, and regular combinators.\bcp{If this is our main technical
  contribution, then it deserves more than a sentence in the intro!  E.g.,
  can we give an example?} This proof will allow justify our
approach for synthesizing quotient lenses so that the programmer can simply
type

@@

We have already encountered the \bcp{Again, formatting is mangled.  Also,
  are we setting lens expressions in typewriter font, or in snas-serif?  I
  don't care, but we need to be consistent.}\verb!project R $\to$ s! QRE
which sends all elements of the language described by the regular expression
$R$, and the
